Mathematical Description

### 2023-12-06 03-post-process/output-plots-children/02-mcmc-diags.Rmd

All convergence diagnostics were calculated using the '[posterior](https://mc-stan.org/posterior/articles/posterior.html)' R package, which implements updated versions (based on rank-normalization) of the classical convergence diagnostics (presented in Vehtari et al. [2021](https://projecteuclid.org/journals/bayesian-analysis/advance-publication/Rank-Normalization-Folding-and-Localization--An-Improved-R%cb%86-for/10.1214/20-BA1221.full)).

* **Rhat** measures convergence of multiple chains to a common target distribution. Values close to 1 are better -- we've chosen these approximate categories for ease of interpretation: 
  * _Ideal_: [1,1.01] 
  * _No Issue_: (1.01,1.05]
  * _Acceptable_: (1.05,1.1]
  * _Potential Concern_: (1.1,1.2]
  * _Definite Concern_: (1.2,100)
* **ESS** measures the number of effective MCMC samples are available to estimate a given summary statistic of the posterior -- here the posterior mean and quantiles of the 95% equal-tailed CRI. Percentages shown represent the fraction of parameters that have fewer than 400 ESS -- Vehtari et al. ([2021](https://projecteuclid.org/journals/bayesian-analysis/advance-publication/Rank-Normalization-Folding-and-Localization--An-Improved-R%cb%86-for/10.1214/20-BA1221.full)) suggests 400 as a good rule of thumb.
* **MCCV** is Monte Carlo Standard Error of a summary statistic divided by the estimated posterior summary statistic, intended to represent a measure of precision for estimating posterior summary statistics -- here the posterior mean and quantiles of the 95% equal-tailed CRI. Percentages shown represent the fraction of parameters that have greater than 5% MCCV, which we have deemed an acceptable level of imprecision.

@@

# which parameters fall in each Rhat bin
rhat_1_params = subset(diags_out, rhat >= 0.8 & rhat < 1.05)$variable
rhat_2_params = subset(diags_out, rhat >= 1.05 & rhat < 1.1)$variable
rhat_3_params = subset(diags_out, rhat >= 1.1 & rhat < 1.5)$variable

@@

# subset out these parameters: makes postpack::diag_plots() much faster
post_sub_1 = post_subset(post, rhat_1_params_draw)
post_sub_2 = post_subset(post, rhat_2_params_draw)
post_sub_3 = post_subset(post, rhat_3_params_draw)

# function to plot a histogram of rhat values
rhat_hist = function(rhat) {
  # par(mgp = c(2,0.35, 0), tcl = -0.15, xaxs = "i", yaxs = "i", mar = c(3,3,1,1))
  mypar(xaxs = "i", yaxs = "i", mfrow = c(1,1))
  breaks = seq(min(rhat), max(rhat), length = 20)
  max_count = max(hist(rhat, breaks = breaks, plot = FALSE)$count)
  hist(rhat, breaks = breaks, col = "grey50", border = "white", ylim = c(0, max_count) * 1.025, main = "")
  axis_labels(xlab = TeX("$\\hat{R}$"), ylab = "Frequency")
  usr = par("usr")
  segments(usr[1], usr[3], usr[2], usr[3], xpd = TRUE, col = par("col.axis"))
  segments(usr[1], usr[3], usr[1], usr[4], xpd = TRUE, col = par("col.axis"))
}
```

### $1 < \hat{R} < 1.05$ {.tabset .tabset-pills}

@@

### $1.05 < \hat{R} < 1.1$ {.tabset .tabset-pills}

@@

### $\hat{R} > 1.1$ {.tabset .tabset-pills}

@@

There were **`r length(rhat_3_params)`** of **`r nrow(diags_out)`** unobserved stochastic nodes in this group. Here is their distribution:

```{r rhat-hist-3, fig.width = 6, fig.height = 4}
rhat_hist(rhat_3)
```

@@

```{r mcmc-diag-plots-3, fig.width = 11, fig.height = 8.5}
diag_plots(post_sub_3, ".", show_diags = "never")
```

>>>>

<details>
<summary>Click to View Diagnostic Descriptions</summary>

* **Rhat** measures convergence of multiple chains to a common target distribution. Values close to 1 are better -- we've chosen these categories for ease of interpretation: 
  * _Ideal_: [1,1.01] 
  * _No Issue_: (1.01,1.05]
  * _Acceptable_: (1.05,1.1]
  * _Potential Concern_: (1.1,1.2]
  * _Definite Concern_: (1.2,100)
* **ESS** measures the number of effective MCMC samples are available to estimate a given summary statistic of the posterior -- here the posterior mean and quantiles of the 95% equal-tailed CRI. Percentages shown represent the fraction of parameters that have fewer than 400 ESS -- Vehtari et al. ([2021](https://projecteuclid.org/journals/bayesian-analysis/advance-publication/Rank-Normalization-Folding-and-Localization--An-Improved-R%cb%86-for/10.1214/20-BA1221.full)) suggests 400 as a good rule of thumb.
* **MCCV** is Monte Carlo Standard Error of a summary statistic divided by the estimated posterior summary statistic, intended to represent a measure of precision for estimating posterior summary statistics -- here the posterior mean and quantiles of the 95% equal-tailed CRI. Percentages shown represent the fraction of parameters that have greater than 5% MCCV, which we have deemed an acceptable level of imprecision.

_All convergence diagnostics were calculated using the '[posterior](https://mc-stan.org/posterior/articles/posterior.html)' R package, which implements updated versions (based on rank-normalization) of the classical convergence diagnostics (presented in Vehtari et al. [2021](https://projecteuclid.org/journals/bayesian-analysis/advance-publication/Rank-Normalization-Folding-and-Localization--An-Improved-R%cb%86-for/10.1214/20-BA1221.full))._

</details>

@@

# which parameters fall in each Rhat bin
rhat_1_params = subset(diags_out, rhat >= 0.8 & rhat < 1.05)$variable
rhat_2_params = subset(diags_out, rhat >= 1.05 & rhat < 1.1)$variable
rhat_3_params = subset(diags_out, rhat >= 1.1)$variable

@@

# subset out these parameters: makes postpack::diag_plots() much faster
post_sub_1 = post_subset(post, rhat_1_params_draw)
post_sub_2 = post_subset(post, rhat_2_params_draw)

if (length(rhat_3_params) > 0) post_sub_3 = post_subset(post, rhat_3_params_draw)

# function to plot a histogram of rhat values
rhat_hist = function(rhat) {
  # par(mgp = c(2,0.35, 0), tcl = -0.15, xaxs = "i", yaxs = "i", mar = c(3,3,1,1))
  mypar(xaxs = "i", yaxs = "i", mfrow = c(1,1))
  breaks = seq(min(rhat), max(rhat), length = 20)
  max_count = max(hist(rhat, breaks = breaks, plot = FALSE)$count)
  hist(rhat, breaks = breaks, col = "grey50", border = "white", ylim = c(0, max_count) * 1.025, main = "")
  axis_labels(xlab = "Rhat", ylab = "Frequency")
  usr = par("usr")
  segments(usr[1], usr[3], usr[2], usr[3], xpd = TRUE, col = par("col.axis"))
  segments(usr[1], usr[3], usr[1], usr[4], xpd = TRUE, col = par("col.axis"))
}
```

### Ideal/No Issue {.tabset .tabset-pills}

@@

### Acceptable {.tabset .tabset-pills}

@@

### Concerning {.tabset .tabset-pills}

@@

There were **`r length(rhat_3_params)`** of **`r nrow(diags_out)`** unobserved stochastic nodes in this group. `r ifelse(nrow(rhat_3_params) > 0, "Here is their distribution:", "")`

```{r rhat-hist-3, fig.width = 6, fig.height = 4, eval = length(rhat_3_params) > 0}
rhat_hist(rhat_3)
```

@@

```{r mcmc-diag-plots-3, fig.width = 11, fig.height = 8.5, eval = length(rhat_3_params) > 0}
diag_plots(post_sub_3, ".", show_diags = "never")
```

### 2023-12-06 03-post-process/output-plots-children/10-rho-params.Rmd

```{r rho_Lphi_Mb_O0-plot}
est = post_summ(post, "rho_Lphi_Ma_O0[2,1]")
kable(t(est), digits = 3, row.names = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("condensed"))
```

@@

```{r rho_Lphi_Rb_Ra-plot, fig.height = rho_mat_fig_h, fig.width = rho_mat_fig_w}
est = post_summ(post, "rho_Lphi_Rb_Ra[2,1]")
kable(t(est), digits = 3, row.names = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("condensed"))
```

>>>>

```{r rho_Lphi_Mb_O0-plot}
est = post_summ(post, "rho_Lphi_Ma_O0[2,1]")
kable(t(est), digits = 3, row.names = FALSE) |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("condensed"))
```

@@

```{r rho_Lphi_Rb_Ra-plot, fig.height = rho_mat_fig_h, fig.width = rho_mat_fig_w}
est = post_summ(post, "rho_Lphi_Rb_Ra[2,1]")
kable(t(est), digits = 3, row.names = FALSE) |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("condensed"))
```

### 2023-12-06 03-post-process/output-plots-children/12-pp-checks.Rmd

# produce the kable
bp_tab %>%
  kbl(digits = 2, col.names = c("Class", "Lifestage", "Population", "Migratory Type", "Origin", "Years w/Data", "Median", "2.5%", "97.5%", "<0.05", ">0.95"),
      align = c("lllllcccccc")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("condensed")) %>%
  add_header_above(c("Data Type Information" = 5, "P-value Summaries" = 4, "% of Extreme P-values" = 2)) %>%
  column_spec(7, background = assign_color(bp_tab[,7])) %>%
  column_spec(8, background = assign_color(bp_tab[,8])) %>%
  column_spec(9, background = assign_color(bp_tab[,9])) %>%
  collapse_rows(1:5) 
```

>>>>

# produce the kable
bp_tab |> 
  kbl(digits = 2, col.names = c("Class", "Lifestage", "Population", "Migratory Type", "Origin", "Years w/Data", "Median", "2.5%", "97.5%", "<0.05", ">0.95"),
      align = c("lllllcccccc")) |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("condensed")) |> 
  add_header_above(c("Data Type Information" = 5, "P-value Summaries" = 4, "% of Extreme P-values" = 2)) |> 
  column_spec(7, background = assign_color(bp_tab[,7])) |> 
  column_spec(8, background = assign_color(bp_tab[,8])) |> 
  column_spec(9, background = assign_color(bp_tab[,9])) |> 
  collapse_rows(1:5) 
```

### 2023-12-06 100-math-description/GR-sslcm-math-description.Rmd

The purpose of developing the state-space life cycle model for Grande Ronde spring Chinook salmon is to estimate posterior distributions of population dynamics parameters (e.g., average survival, rearing capacity, coefficients of relationships linking life stages, etc.).
These posterior distributions can ultimately be sampled from to parameterize simulation models that evaluate habitat restoration and climate change scenarios grounded in empirical monitoring data.
The model tracks the cohort-specific abundance of fish through various life stages and links these "state variables" over time across cohorts via production relationships (e.g., spawners produce eggs) or within cohorts via transition probabilities (e.g., parr become smolt after surviving overwinter).
The model integrates many different data sources and fits to them simultaneously in a single joint likelihood; these data sources include rotary screw trap passage estimates, a variety of PIT tag-derived survival estimates, mean length data, hatchery smolt releases, adult return abundance, and the composition of adult returns by age and origin; other data sources that are used but are not fitted to include hatchery smolt release abundances, harvest information, fecundity estiamtes, and numbers of returning adults removed for hatchery operations.
Four spawning populations within the Grande Ronde basin are simultaneously analyzed: Catherine Creek (CAT), Lostine River (LOS), Minam River (MIN) and the upper Grande Ronde River (UGR; `r highlight("FIGURE REFERENCE: map")`), and where relevant, the model stratifies life stages by juvenile migratory strategy (fall vs. spring migrants), rearing origin type (spawned in hatchery vs. natural setting), and adult return age (total age of 3, 4, or 5 years).

@@

# build the table
my_kable(x = tab[,1:3], caption = index_tab_cap) %>%
  kable_styling(full_width = FALSE, latex_options = c("HOLD")) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "50px") %>%
  column_spec(2, width = "50px") %>%
  column_spec(3, width = "332px") %>%
  collapse_rows(1, valign = "bottom", latex_hline = "major") %>%
  footnote(number = footnote_key, threeparttable = TRUE, escape = FALSE)
```

@@

my_kable(x = tab[,2:5], caption = symbol_tab_cap, longtable = TRUE) %>%
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) %>%
  row_spec(0, bold = TRUE) %>%
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Ocean Juvenile", min(which(tab$Phase == 2)), max(which(tab$Phase == 2)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) %>%
  column_spec(1, width = "25px") %>%
  column_spec(2, width = "50px") %>%
  column_spec(3, width = "50px") %>%
  column_spec(4, width = "300px") %>%
  collapse_rows(1, valign = "bottom", latex_hline = "full")
```

@@

my_kable(x = tab[,2:5], caption = param_tab_cap, longtable = TRUE) %>%
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) %>%
  row_spec(0, bold = TRUE) %>%
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Ocean Juvenile", min(which(tab$Phase == 2)), max(which(tab$Phase == 2)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) %>%
  column_spec(1, width = "50px") %>%
  column_spec(2, width = "65px") %>%
  column_spec(3, width = "20px") %>%
  column_spec(4, width = "290px") %>%
  row_spec(1:(nrow(tab)-1), hline_after = TRUE) %>%
  footnote(number = footnote_key, number_title = "Prior Distributions:", title_format = c("italic", "underline"), escape = FALSE, threeparttable = TRUE)
```

@@

my_kable(x = tab[,2:5], caption = data_tab_cap, longtable = TRUE) %>%
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) %>%
  row_spec(0, bold = TRUE) %>%
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) %>%
  column_spec(1, width = "50px") %>%
  column_spec(2, width = "70px") %>%
  column_spec(3, width = "20px") %>%
  column_spec(4, width = "295px") %>%
  collapse_rows(1, valign = "bottom", latex_hline = "full")
```

@@

my_kable(x = tab[,2:4], caption = equation_tab_cap, longtable = TRUE, align = "lrl") %>%
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) %>%
  row_spec(0, bold = TRUE) %>%
  pack_rows("Generic", min(which(tab$Phase == 0)), max(which(tab$Phase == 0)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Ocean Juvenile", min(which(tab$Phase == 2)), max(which(tab$Phase == 2)), bold = TRUE, hline_after = TRUE) %>%
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) %>%
  column_spec(1, width = "55px") %>%
  column_spec(2, width = "15px") %>%
  column_spec(3, width = "365px") %>%
  collapse_rows(1, valign = "bottom", latex_hline = "major")
```

@@

my_kable(x = tab, align = "cccc", col.names = c("Return Year", "Age-3", "Age-4", "Age-5"), 
         caption = year_tab_caption, row.names = FALSE) %>%
  kable_styling(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Brood Year of Returning Adults by Age" = 3), bold = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:4, width = "50px")
```

>>>>

The purpose of developing the state-space life cycle model for Grande Ronde spring Chinook salmon is to estimate posterior distributions of population dynamics parameters (e.g., average survival, rearing capacity, coefficients of relationships linking life stages, etc.).
These posterior distributions can ultimately be sampled from to parameterize simulation models that evaluate habitat restoration and climate change scenarios grounded in empirical monitoring data.
The model tracks the cohort-specific abundance of fish through various life stages and links these "state variables" over time across cohorts via production relationships (e.g., spawners produce eggs) or within cohorts via transition probabilities (e.g., parr become smolt after surviving overwinter).
The model integrates many different data sources and fits to them simultaneously in a single joint likelihood; these data sources include rotary screw trap passage estimates, a variety of PIT tag-derived survival estimates, mean length data, hatchery smolt releases, adult return abundance, and the composition of adult returns by age and origin; other data sources that are used but are not fitted to include hatchery smolt release abundances, harvest information, fecundity estimates, and numbers of returning adults removed for hatchery operations.
Four spawning populations within the Grande Ronde basin are simultaneously analyzed: Catherine Creek (CAT), Lostine River (LOS), Minam River (MIN) and the upper Grande Ronde River (UGR; `r highlight("FIGURE REFERENCE: map")`), and where relevant, the model stratifies life stages by juvenile migratory strategy (fall vs. spring migrants), rearing origin type (spawned in hatchery vs. natural setting), and adult return age (total age of 3, 4, or 5 years).

@@

# build the table
my_kable(x = tab[,1:3], caption = index_tab_cap) |> 
  kable_styling(full_width = FALSE, latex_options = c("HOLD")) |> 
  row_spec(0, bold = TRUE) |> 
  column_spec(1, width = "50px") |> 
  column_spec(2, width = "50px") |> 
  column_spec(3, width = "332px") |> 
  collapse_rows(1, valign = "bottom", latex_hline = "major") |> 
  footnote(number = footnote_key, threeparttable = TRUE, escape = FALSE)
```

@@

my_kable(x = tab[,2:5], caption = symbol_tab_cap, longtable = TRUE) |> 
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) |> 
  row_spec(0, bold = TRUE) |> 
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Ocean Juvenile", min(which(tab$Phase == 2)), max(which(tab$Phase == 2)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) |> 
  column_spec(1, width = "25px") |> 
  column_spec(2, width = "50px") |> 
  column_spec(3, width = "50px") |> 
  column_spec(4, width = "300px") |> 
  collapse_rows(1, valign = "bottom", latex_hline = "full")
```

@@

my_kable(x = tab[,2:5], caption = param_tab_cap, longtable = TRUE) |> 
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) |> 
  row_spec(0, bold = TRUE) |> 
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Ocean Juvenile", min(which(tab$Phase == 2)), max(which(tab$Phase == 2)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) |> 
  column_spec(1, width = "50px") |> 
  column_spec(2, width = "65px") |> 
  column_spec(3, width = "20px") |> 
  column_spec(4, width = "290px") |> 
  row_spec(1:(nrow(tab)-1), hline_after = TRUE) |> 
  footnote(number = footnote_key, number_title = "Prior Distributions:", title_format = c("italic", "underline"), escape = FALSE, threeparttable = TRUE)
```

@@

my_kable(x = tab[,2:5], caption = data_tab_cap, longtable = TRUE) |> 
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) |> 
  row_spec(0, bold = TRUE) |> 
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) |> 
  column_spec(1, width = "50px") |> 
  column_spec(2, width = "70px") |> 
  column_spec(3, width = "20px") |> 
  column_spec(4, width = "295px") |> 
  collapse_rows(1, valign = "bottom", latex_hline = "full")
```

@@

my_kable(x = tab[,2:4], caption = equation_tab_cap, longtable = TRUE, align = "lrl") |> 
  kable_styling(full_width = FALSE, latex_options = c("HOLD", "repeat_header")
                # repeat_header_continued = "\\textit{(Continued on next page...)}"
                ) |> 
  row_spec(0, bold = TRUE) |> 
  pack_rows("Generic", min(which(tab$Phase == 0)), max(which(tab$Phase == 0)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Freshwater Juvenile", min(which(tab$Phase == 1)), max(which(tab$Phase == 1)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Ocean Juvenile", min(which(tab$Phase == 2)), max(which(tab$Phase == 2)), bold = TRUE, hline_after = TRUE) |> 
  pack_rows("Freshwater Adult", min(which(tab$Phase == 3)), max(which(tab$Phase == 3)), bold = TRUE, hline_after = TRUE) |> 
  column_spec(1, width = "55px") |> 
  column_spec(2, width = "15px") |> 
  column_spec(3, width = "365px") |> 
  collapse_rows(1, valign = "bottom", latex_hline = "major")
```

@@

my_kable(x = tab, align = "cccc", col.names = c("Return Year", "Age-3", "Age-4", "Age-5"), 
         caption = year_tab_caption, row.names = FALSE) |> 
  kable_styling(full_width = FALSE) |> 
  add_header_above(c(" " = 1, "Brood Year of Returning Adults by Age" = 3), bold = TRUE) |> 
  row_spec(0, bold = TRUE) |> 
  column_spec(1, bold = TRUE) |> 
  column_spec(2:4, width = "50px")
```
